\huge \textbf{Unshuffling Permutations} \vspace{1.5em}

### 2016-04-08 slides/slides.tex
Modifs

\DeclareMathOperator{\QQ}{\mathbb{Q}}
\DeclareMathOperator{\STD}{\mathrm{std}}
\DeclareMathOperator{\SHUFFLE}{\bullet}
\DeclareMathOperator{\La}{\mathtt{a}}
\DeclareMathOperator{\Lb}{\mathtt{b}}

@@

\title[Unshuffling Permutations]{%
  \textbf{Unshuffling Permutations}
}

\subtitle{
    \textbf{Samuele Giraudo}\quad\textbf{St\'ephane~Vialette}\\
    Laboratoire d'Informatique Gaspard-Monge\\
    Université Paris-Est Marne-la-Vallée \\
    UMR CNRS 8049 \\
    ~\\
    \includegraphics[width=1.cm,height=1.25cm,keepaspectratio]{logos/CNRSfr}
    \qquad
    \includegraphics[width=1.cm,height=1.25cm,keepaspectratio]{logos/ecole_ponts_RVB300.jpg}
    \qquad
    \includegraphics[width=1.5cm,height=2cm,keepaspectratio]{logos/LIGM}
    \qquad
    \raisebox{.25cm}{\includegraphics[width=1.cm,height=1.25cm,keepaspectratio]{logos/LogoEsieeParis}}
    \qquad
    \raisebox{.25cm}{\includegraphics[width=1.cm,height=1.25cm,keepaspectratio]{logos/UPEM_LOGO_EDITION}}
}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{frame} \frametitle{Shuffling words}
\begin{block}{Definition}
    The \alert{shuffle product} $\shuffle$ on words of $A^*$ is defined
    inductively by
    \begin{align*}
        u \shuffle \epsilon &= \epsilon \shuffle u = \{u\} \\
        (ua \shuffle vb) &= (ua \shuffle v)b \cup (u \shuffle vb)a
    \end{align*}
\end{block}
\medskip

\uncover<2->{%
To take into account multiplicities, we consider $\shuffle$ as a linear
product
\begin{equation*}
    \QQ[A^*] \otimes \QQ[A^*] \to \QQ[A^*]
\end{equation*}
on $\QQ[A^*]$, the $\QQ$-linear span of words defined by
\begin{align*}
    u \shuffle \epsilon & = \epsilon \shuffle u = u \\
    (ua \shuffle vb) &= (ua \shuffle v)b + (u \shuffle vb)a
\end{align*}}

\uncover<3->{%
\begin{block}{Example}
    \vspace{-1em}
    \begin{equation*}\begin{split}
        \textcolor{RoyalBlue}{\La\Lb} \shuffle \textcolor{BrickRed}{\Lb\La}
        & =
        \textcolor{RoyalBlue}{\La\Lb}\textcolor{BrickRed}{\Lb\La}
        + \textcolor{RoyalBlue}{\La}\textcolor{BrickRed}{\Lb}%
        \textcolor{RoyalBlue}{\Lb}\textcolor{BrickRed}{\La}
        + \textcolor{RoyalBlue}{\La}\textcolor{BrickRed}{\Lb\La}%
        \textcolor{RoyalBlue}{\Lb}
        + \textcolor{BrickRed}{\Lb}\textcolor{RoyalBlue}{\La\Lb}%
        \textcolor{BrickRed}{\La}
        + \textcolor{BrickRed}{\Lb}\textcolor{RoyalBlue}{\La}%
        \textcolor{BrickRed}{\La}\textcolor{RoyalBlue}{\Lb}
        + \textcolor{BrickRed}{\Lb\La}\textcolor{RoyalBlue}{\La\Lb} \\
        &
        \uncover<4->{=
        2 \La\Lb\Lb\La + \La\Lb\La\Lb + \Lb\La\Lb\La + 2 \Lb\La\La\Lb}
    \end{split}\end{equation*}
    \vspace{-1.5em}
\end{block}}
\end{frame}

@@

Given three words $u$, $v_1$, and $v_2$, deciding if $u$ is a
shuffle of $v_1$ and $v_2$ can be done in
\begin{equation*}
O\left(\frac{|u|^2}{\log(|u|)}\right)
\end{equation*}
time \BIB{van Leeuwen, Nivat}{1982}.
\bigskip

Given a word $u$, deciding if there is a word $v$ such that $u$ is
in $v \shuffle v$ is \NPC{} \BIB{Rizzi, Vialette}{2013}
\BIB{Buss, Soltys}{2014}.
\end{frame}

@@

\begin{block}{Example}
    The word
    \begin{math}
        u := \textcolor{RoyalBlue}{3}\textcolor{BrickRed}{3}
        \textcolor{RoyalBlue}{12}\textcolor{BrickRed}{1}
        \textcolor{RoyalBlue}{2}\textcolor{BrickRed}{22}
    \end{math}
    is a square since $u$ can be obtained by shuffling $3122$ with itself.
\end{block}
\medskip

The first numbers of square binary words of length $2n$ are
\begin{equation*}
    1, 2, 6, 22, 82, 320, 1268, 5102, 20632, 83972.
\end{equation*}

\begin{block}{Open problem}
    Enumeration of square (binary) words.
\end{block}
\end{frame}

@@

\begin{block}{Example}
    The word $21212233$ admits the perfect matching \vspace{-1em}
    \begin{equation*}
    \begin{split}
    \begin{tikzpicture}[xscale=.5,yscale=.3,inner sep=1pt,node distance=1.5cm]
        \tikzstyle{Vertex}=[draw,circle,minimum size=.35cm,font=\footnotesize]
        \draw node[Vertex,fill=RoyalBlue!30](U01){$(2, 1)$};
        \draw node[Vertex,fill=RoyalBlue!30,right of=U01](U02) {$(1, 2)$};
        \draw node[Vertex,fill=RoyalBlue!30,right of=U02](U03){$(2, 3)$};
        \draw node [Vertex,fill=BrickRed!30,right of=U03](U04) {$(1, 4)$};
        \draw node [Vertex,fill=BrickRed!30,right of=U04](U05) {$(2, 5)$};
        \draw node [Vertex,fill=BrickRed!30,right of=U05](U06){$(2, 6)$};
        \draw node [Vertex,fill=RoyalBlue!30,right of=U06](U07){$(3, 7)$};
        \draw node [Vertex,fill=BrickRed!30,right of=U07](U08){$(3, 8)$};
        \draw [thick,>=latex']
        (U01.north) .. controls ($ (U01.north) + (0,2.5) $)
            and ($ (U06.north) + (0,2.5) $) .. (U06.north);
        \draw [thick,>=latex']
        (U02.north) .. controls ($ (U02.north) + (0,1.5) $)
            and ($ (U04.north) + (0,1.5) $) .. (U04.north);
        \draw [thick,>=latex']
        (U03.south) .. controls ($ (U03.south) + (0,-1.5) $)
            and ($ (U05.south) + (0,-1.5) $) .. (U05.south);
        \draw [thick,>=latex']
        (U07.north) .. controls ($ (U07.north) + (0,1.5) $)
            and ($ (U08.north) + (0,1.5) $) .. (U08.north);
    \end{tikzpicture}
    \end{split}\,.
    \end{equation*}
    \vspace{-1.5em}
\end{block}

\begin{block}{Definition}
    A perfect matching $(V, E)$ of a word $u$ is \alert{inclusion-free}
    if there are no edges $(u_i, i) \-- (u_j, j)$ and
    $(u_k, k) \-- (u_\ell, \ell)$ of $E$ such that $i < k < \ell < j$.
\end{block}
\end{frame}

@@

\begin{block}{Proposition \BIB{Rizzi, Vialette}{2013}}
    A word $u$ is a square iff $u$ admits an inclusion-free perfect
    matching.
\end{block}
%\medskip

\begin{block}{Example}
    The square word
    \begin{math}
        \textcolor{RoyalBlue}{3}\textcolor{BrickRed}{3}
        \textcolor{RoyalBlue}{12}\textcolor{BrickRed}{1}
        \textcolor{RoyalBlue}{2}\textcolor{BrickRed}{22}
    \end{math}
    admits the perfect matching \vspace{-1.5em}
    \begin{equation*}
        \begin{split}
        \begin{tikzpicture}[xscale=.5,yscale=.2,inner sep=1pt,node distance=1.5cm]
            \tikzstyle{Vertex}=[draw,circle,minimum size=.35cm,font=\footnotesize]
            \draw node[Vertex,fill=RoyalBlue!30](U01){$(3, 1)$};
            \draw node[Vertex,fill=BrickRed!30,right of=U01](U02) {$(3, 2)$};
            \draw node[Vertex,fill=RoyalBlue!30,right of=U02](U03){$(1, 3)$};
            \draw node[Vertex,fill=RoyalBlue!30,right of=U03](U04) {$(2, 4)$};
            \draw node[Vertex,fill=BrickRed!30,right of=U04](U05) {$(1, 5)$};
            \draw node[Vertex,fill=RoyalBlue!30,right of=U05](U06){$(2, 6)$};
            \draw node[Vertex,fill=BrickRed!30,right of=U06](U07){$(2, 7)$};
            \draw node[Vertex,fill=BrickRed!30,right of=U07](U08){$(2, 8)$};
            \draw [thick,>=latex']
            (U01.north) .. controls ($ (U01.north) + (0,2.25) $)
                and ($ (U02.north) + (0,2.25) $) .. (U02.north);
            \draw [thick,>=latex']
            (U03.north) .. controls ($ (U03.north) + (0,2.25) $)
                and ($ (U05.north) + (0,2.25) $) .. (U05.north);
            \draw [thick,>=latex']
            (U04.north) .. controls ($ (U04.north) + (0,4.5) $)
                and ($ (U07.north) + (0,4.5) $) .. (U07.north);
            \draw [thick,>=latex']
            (U05.south) .. controls ($ (U05.south) + (0,-2.5) $)
                and ($ (U08.south) + (0,-2.5) $) .. (U08.south);
            \draw [thick,>=latex']
            (U06.north) .. controls ($ (U06.north) + (0,2.5) $)
                and ($ (U08.north) + (0,2.5) $) .. (U08.north);
        \end{tikzpicture}
        \end{split}
    \end{equation*}

    \vspace{-1.5em}
    which is inclusion-free.
\end{block}
%\medskip

\begin{block}{Example}
    The word $1221$ is not a square. Its only perfect matching\vspace{-1.5em}
    \begin{equation*}
        \begin{split}
        \begin{tikzpicture}[xscale=.5,yscale=.3,inner sep=1pt,node distance=1.5cm]
            \tikzstyle{Vertex}=[draw,circle,minimum size=.35cm,font=\footnotesize]
            \draw node[Vertex,fill=RoyalBlue!30](U01){$(1, 1)$};
            \draw node[Vertex,fill=RoyalBlue!30,right of=U01](U02) {$(2, 2)$};
            \draw node[Vertex,fill=BrickRed!30,right of=U02](U03){$(2, 3)$};
            \draw node [Vertex,fill=BrickRed!30,right of=U03](U04) {$(1, 4)$};
            \draw [thick,>=latex']
            (U01.north) .. controls ($ (U01.north) + (0,2.5) $)
                and ($ (U04.north) + (0,2.5) $) .. (U04.north);
            \draw [thick,>=latex']
            (U02.north) .. controls ($ (U02.north) + (0,1.5) $)
                and ($ (U03.north) + (0,1.5) $) .. (U03.north);
        \end{tikzpicture}\vspace{-1.5em}
        \end{split}
    \end{equation*}

    \vspace{-1.5em}
    is not inclusion-free.
\end{block}
\end{frame}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%o%%%%%%%%%%%%%%%%%%
\begin{frame} \frametitle{Main motivations}
Let $S_n$ be the set of permutations of size $n$.
\medskip

There are several products on permutations, analogs of the shuffle
product of words. Among these, there are
\smallskip

@@

    \item the supershuffle \BIB{Vargas}{2014}.
\end{itemize}
\medskip

\begin{block}{Main questions}
    \begin{itemize}
        \item Combinatorial properties of ``square permutations''
        w.r.t. the supershuffle?
        \medskip

        \item Complexity of the recognition of ``square permutations''?
    \end{itemize}
\end{block}
\end{frame}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{frame} \frametitle{Example: unshuffling binary words}
Let $B$ the set of all binary words and let us consider the coproduct
\begin{equation*}
    \Delta : \QQ[B] \to \QQ[B] \otimes \QQ[B]
\end{equation*}
on binary words defined by
\begin{equation*}
    \Delta(u) :=
    \sum_{P \sqcup Q = \{1, \dots, |u|\}}
    u_{|P} \otimes u_{|Q}.
\end{equation*}

\begin{block}{Example}
\begin{equation*}
    \Delta(100) =
    \epsilon \otimes 100 + 1 \otimes 00
    + 2 (0 \otimes 10) + 2 (10 \otimes 0) + 00 \otimes 1
    + 100 \otimes \epsilon
\end{equation*}
\end{block}
\end{frame}

@@

\begin{block}{Example}
Since
\begin{equation*}
    \textcolor{RoyalBlue}{\Delta(100)} =
    \epsilon \otimes 100 + 1 \otimes 00
    + \textcolor{Orange}{2} (\textcolor{BrickRed}{0 \otimes 10})
    + 2 (10 \otimes 0) + 00 \otimes 1
    + 100 \otimes \epsilon,
\end{equation*}
we observe that $\textcolor{BrickRed}{0 \otimes 10} $ has multiplicity
$\textcolor{Orange}{2}$ in $\textcolor{RoyalBlue}{\Delta(100)}$.
\medskip

By duality, this implies that $\textcolor{RoyalBlue}{100}$ has
multiplicity $\textcolor{Orange}{2}$ in
$\textcolor{BrickRed}{0 \shuffle 10}$. Indeed,
\begin{equation*}
    \textcolor{BrickRed}{0 \shuffle 10} =
    010 + \textcolor{Orange}{2} (\textcolor{RoyalBlue}{100}).
\end{equation*}
\end{block}
\end{frame}

@@

\begin{block}{Example}
\begin{equation*}
    \STD(82194) = 42153
\end{equation*}
\end{block}
\medskip

\begin{block}{Definition}
    The \alert{unshuffling coproduct of permutations} is the
    coproduct on $\QQ[S]$ defined by
    \begin{equation*}
        \Delta(\pi) =
        \sum_{P \sqcup Q = \{1, \dots, |\pi|\}}
        \STD\left(\pi_{|P}\right) \otimes \STD\left(\pi_{|Q}\right).
    \end{equation*}
\end{block}

\begin{block}{Example}
\begin{equation*}
    \Delta(213) =
    \epsilon \otimes 213 + 2 (1 \otimes 12) +
    1 \otimes 21 + 2 (12 \otimes 1) + 21 \otimes 1
    + 213 \otimes \epsilon
\end{equation*}
\end{block}
\end{frame}

@@

\begin{block}{Proposition}
The permutations appearing in $\sigma \SHUFFLE \nu$ are exactly the
one obtained by shuffling two words $u$ and $v$ respectively
order-isomorphic to $\sigma$ and $\nu$.
\end{block}
\medskip

\begin{block}{Example}
\begin{equation*}\begin{split}
    12 \SHUFFLE 21 & =
    1243 + 1324 + 2 (1342) + 2 (1423) + 3 (1432) +
    2134 + 2 (2314) \\[.5em]
    & + 3 (2341) + 2413 + 2 (2431) + 2 (3124) + 3142 +
    3 (3214) + 2 (3241) \\[.5em]
    & + 3421 + 3 (4123) + 2 (4132) + 2 (4213) + 4231 + 4312
\end{split}\end{equation*}
\end{block}
\end{frame}

@@

\begin{block}{Proposition}
    A permutation $\pi$ is a square iff $\pi$ can be obtained by
    shuffling two order-isomorphic words.
\end{block}
\medskip

\begin{block}{Example}
    The permutation
    \begin{math}
        \pi :=
        \textcolor{RoyalBlue}{25}\textcolor{BrickRed}{16}
        \textcolor{RoyalBlue}{7}\textcolor{BrickRed}{83}
        \textcolor{RoyalBlue}{4}
    \end{math}
    is square since $\pi$ can be obtained by shuffling
    $\textcolor{BrickRed}{1683}$ and $\textcolor{RoyalBlue}{2574}$
    and
    \begin{math}
        \STD(\textcolor{BrickRed}{1683})
        = \STD(\textcolor{RoyalBlue}{2574})
        = 1342
    \end{math}.
\end{block}
\end{frame}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{frame} \frametitle{Some properties of square permutations}
The first numbers of square permutations of size $2$ are
\begin{equation*}
    1, 2, 20, 504, 21032, 1293418.
\end{equation*}
%\medskip

\begin{block}{Proposition}
    Let $\pi$ be a square permutation and $\sigma$ be a square root of
    $\pi$. Then,
    \begin{enumerate}
        \item $\widetilde{\pi}$ is a square and
        $\widetilde{\sigma}$ is one of its square roots;
        \smallskip

@@

        \item $\pi^{-1}$ is a square and $\sigma^{-1}$ is one of
        its square roots.
    \end{enumerate}
\end{block}
\medskip

>>>>

\DeclareMathOperator{\QQ}{\mathbb{Q}}
\DeclareMathOperator{\STD}{\mathrm{std}}
\DeclareMathOperator{\SHUFFLE}{\bullet}
\DeclareMathOperator{\La}{\mathtt{a}}
\DeclareMathOperator{\Lb}{\mathtt{b}}
\DeclareMathOperator{\Lc}{\mathtt{c}}
\DeclareMathOperator{\Ld}{\mathtt{d}}

@@

\title[Unshuffling Permutations]{%
  \huge \textbf{Unshuffling Permutations}
  \vspace{1.5em}
}

\subtitle{
    \textbf{\underline{Samuele Giraudo}}\qquad\textbf{St\'ephane~Vialette}\\[1em]
    {\footnotesize Laboratoire d'Informatique Gaspard-Monge\\
    Université Paris-Est Marne-la-Vallée \\
    UMR CNRS 8049 \\[2em]}
    \includegraphics[scale=.16]{logos/CNRSfr}
    \qquad
    \includegraphics[scale=.10]{logos/ecole_ponts_RVB300.jpg}
    \qquad
    \includegraphics[scale=.26]{logos/LIGM}
    \qquad
    \raisebox{.25cm}{\includegraphics[width=1.cm,height=1.25cm,keepaspectratio]{logos/LogoEsieeParis}}
    \qquad
    \raisebox{.25cm}{\includegraphics[width=1.cm,height=1.25cm,keepaspectratio]{logos/UPEM_LOGO_EDITION}}
}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{frame} \frametitle{Shuffling words}
\begin{block}{Definition}
    The \alert{shuffle product} $\shuffle$ on words of $A^*$ is defined
    recursively by \vspace{-.5em}
    \begin{align*}
        u \shuffle \epsilon & := \{u\} =: \epsilon \shuffle u, \\
        (ua \shuffle vb) & := (ua \shuffle v)b \cup (u \shuffle vb)a.
    \end{align*}
\end{block}
\medskip

\uncover<2->{%
To take into account multiplicities, we consider $\shuffle$ as a linear
product
\begin{equation*}
    \QQ[A^*] \otimes \QQ[A^*] \to \QQ[A^*]
\end{equation*}
on $\QQ[A^*]$, the $\QQ$-linear span of words defined by
\begin{align*}
    u \shuffle \epsilon & := u =: \epsilon \shuffle u, \\
    (ua \shuffle vb) & := (ua \shuffle v)b + (u \shuffle vb)a.
\end{align*}}
%
\uncover<3->{%
\begin{block}{Example}
    \vspace{-1em}
    \begin{equation*}\begin{split}
        \textcolor{RoyalBlue}{\La\Lb} \shuffle \textcolor{BrickRed}{\Lb\La}
        & =
        \textcolor{RoyalBlue}{\La\Lb}\textcolor{BrickRed}{\Lb\La}
        + \textcolor{RoyalBlue}{\La}\textcolor{BrickRed}{\Lb}%
        \textcolor{RoyalBlue}{\Lb}\textcolor{BrickRed}{\La}
        + \textcolor{RoyalBlue}{\La}\textcolor{BrickRed}{\Lb\La}%
        \textcolor{RoyalBlue}{\Lb}
        + \textcolor{BrickRed}{\Lb}\textcolor{RoyalBlue}{\La\Lb}%
        \textcolor{BrickRed}{\La}
        + \textcolor{BrickRed}{\Lb}\textcolor{RoyalBlue}{\La}%
        \textcolor{BrickRed}{\La}\textcolor{RoyalBlue}{\Lb}
        + \textcolor{BrickRed}{\Lb\La}\textcolor{RoyalBlue}{\La\Lb} \\
        &
        \uncover<4->{=
        2 \La\Lb\Lb\La + \La\Lb\La\Lb + \Lb\La\Lb\La + 2 \Lb\La\La\Lb}
    \end{split}\end{equation*}
    \vspace{-1.5em}
\end{block}}
\end{frame}

@@

\uncover<2->{%
Given three words $u$, $v_1$, and $v_2$, deciding if $u$ is in
$v_1 \shuffle v_2$ can be done in
\begin{equation*}
O\left(\frac{|u|^2}{\log(|u|)}\right)
\end{equation*}
time \BIB{van Leeuwen, Nivat}{1982}.
\bigskip}

\uncover<3->{%
Given a word $u$, deciding if there is a word $v$ such that $u$ is
in $v \shuffle v$ is \NPC{} \BIB{Rizzi, Vialette}{2013}
\BIB{Buss, Soltys}{2014}.}
\end{frame}

@@

\uncover<2->{%
\begin{block}{Example}
    The word
    \begin{math}
        u := \textcolor{RoyalBlue}{\Lc}\textcolor{BrickRed}{\Lc}
        \textcolor{RoyalBlue}{\La\Lb}\textcolor{BrickRed}{\La}
        \textcolor{RoyalBlue}{\Lb}\textcolor{BrickRed}{\Lb\Lb}
    \end{math}
    is a square since $u$ can be obtained by shuffling $\Lc\La\Lb\Lb$
    with itself.
\end{block}
\medskip}

\uncover<3->{%
The first numbers of square binary words of length $2n$ are
\begin{equation*}
    1, 2, 6, 22, 82, 320, 1268, 5102, 20632, 83972.
\end{equation*}}

\uncover<4->{%
\begin{block}{Open problem}
    Enumeration of square (binary) words.
\end{block}}
\end{frame}

@@

\uncover<2->{%
\begin{block}{Example}
    The word $\Lb\La\Lb\La\Lb\Lb\Lc\Lc$ admits the perfect matching \vspace{-1em}
    \begin{equation*}
    \begin{split}
    \begin{tikzpicture}[xscale=.5,yscale=.3,inner sep=1pt,node distance=1.5cm]
        \tikzstyle{Vertex}=[draw,circle,minimum size=.35cm,font=\footnotesize]
        \draw node[Vertex,fill=RoyalBlue!30](U01){$(\Lb, 1)$};
        \draw node[Vertex,fill=RoyalBlue!30,right of=U01](U02) {$(\La, 2)$};
        \draw node[Vertex,fill=RoyalBlue!30,right of=U02](U03){$(\Lb, 3)$};
        \draw node [Vertex,fill=BrickRed!30,right of=U03](U04) {$(\La, 4)$};
        \draw node [Vertex,fill=BrickRed!30,right of=U04](U05) {$(\Lb, 5)$};
        \draw node [Vertex,fill=BrickRed!30,right of=U05](U06){$(\Lb, 6)$};
        \draw node [Vertex,fill=RoyalBlue!30,right of=U06](U07){$(\Lc, 7)$};
        \draw node [Vertex,fill=BrickRed!30,right of=U07](U08){$(\Lc, 8)$};
        \draw [thick,>=latex']
        (U01.north) .. controls ($ (U01.north) + (0,2.5) $)
            and ($ (U06.north) + (0,2.5) $) .. (U06.north);
        \draw [thick,>=latex']
        (U02.north) .. controls ($ (U02.north) + (0,1.5) $)
            and ($ (U04.north) + (0,1.5) $) .. (U04.north);
        \draw [thick,>=latex']
        (U03.south) .. controls ($ (U03.south) + (0,-1.5) $)
            and ($ (U05.south) + (0,-1.5) $) .. (U05.south);
        \draw [thick,>=latex']
        (U07.north) .. controls ($ (U07.north) + (0,1.5) $)
            and ($ (U08.north) + (0,1.5) $) .. (U08.north);
    \end{tikzpicture}
    \end{split}\,.
    \end{equation*}
    \vspace{-1.5em}
\end{block}}

\uncover<3->{%
\begin{block}{Definition}
    A perfect matching $(V, E)$ of a word $u$ is \alert{inclusion-free}
    if there are no edges $(u_i, i) \-- (u_j, j)$ and
    $(u_k, k) \-- (u_\ell, \ell)$ of $E$ such that $i < k < \ell < j$.
\end{block}}
\end{frame}

@@

\begin{block}{Proposition \BIB{Rizzi, Vialette}{2013}}
    A word $u$ is a square iff $u$ admits an inclusion-free perfect
    matching.
\end{block}

\uncover<2->{%
\begin{block}{Example}
    The word
    \begin{math}
        \textcolor{RoyalBlue}{\Lc}\textcolor{BrickRed}{\Lc}
        \textcolor{RoyalBlue}{\La\Lb}\textcolor{BrickRed}{\La}
        \textcolor{RoyalBlue}{\Lb}\textcolor{BrickRed}{\Lb\Lb}
    \end{math}
    admits the perfect matching \vspace{-1.25em}
    \begin{equation*}
        \begin{split}
        \begin{tikzpicture}[xscale=.5,yscale=.16,inner sep=1pt,node distance=1.5cm]
            \tikzstyle{Vertex}=[draw,circle,minimum size=.35cm,font=\footnotesize]
            \draw node[Vertex,fill=RoyalBlue!30](U01){$(\Lc, 1)$};
            \draw node[Vertex,fill=BrickRed!30,right of=U01](U02) {$(\Lc, 2)$};
            \draw node[Vertex,fill=RoyalBlue!30,right of=U02](U03){$(\La, 3)$};
            \draw node[Vertex,fill=RoyalBlue!30,right of=U03](U04) {$(\Lb, 4)$};
            \draw node[Vertex,fill=BrickRed!30,right of=U04](U05) {$(\La, 5)$};
            \draw node[Vertex,fill=RoyalBlue!30,right of=U05](U06){$(\Lb, 6)$};
            \draw node[Vertex,fill=BrickRed!30,right of=U06](U07){$(\Lb, 7)$};
            \draw node[Vertex,fill=BrickRed!30,right of=U07](U08){$(\Lb, 8)$};
            \draw [thick,>=latex']
            (U01.north) .. controls ($ (U01.north) + (0,2.25) $)
                and ($ (U02.north) + (0,2.25) $) .. (U02.north);
            \draw [thick,>=latex']
            (U03.north) .. controls ($ (U03.north) + (0,2.25) $)
                and ($ (U05.north) + (0,2.25) $) .. (U05.north);
            \draw [thick,>=latex']
            (U04.north) .. controls ($ (U04.north) + (0,4.5) $)
                and ($ (U07.north) + (0,4.5) $) .. (U07.north);
            \draw [thick,>=latex']
            (U05.south) .. controls ($ (U05.south) + (0,-2.5) $)
                and ($ (U08.south) + (0,-2.5) $) .. (U08.south);
            \draw [thick,>=latex']
            (U06.north) .. controls ($ (U06.north) + (0,2.5) $)
                and ($ (U08.north) + (0,2.5) $) .. (U08.north);
        \end{tikzpicture}
        \end{split}
    \end{equation*}

    \vspace{-1.25em}
    which is inclusion-free and hence is a square.
\end{block}}

\uncover<3->{%
\begin{block}{Example}
    The word $\La\Lb\Lb\La$ is not a square. Its only perfect matching\vspace{-1.25em}
    \begin{equation*}
        \begin{split}
        \begin{tikzpicture}[xscale=.5,yscale=.2,inner sep=1pt,node distance=1.5cm]
            \tikzstyle{Vertex}=[draw,circle,minimum size=.35cm,font=\footnotesize]
            \draw node[Vertex,fill=RoyalBlue!30](U01){$(\La, 1)$};
            \draw node[Vertex,fill=RoyalBlue!30,right of=U01](U02) {$(\Lb, 2)$};
            \draw node[Vertex,fill=BrickRed!30,right of=U02](U03){$(\Lb, 3)$};
            \draw node [Vertex,fill=BrickRed!30,right of=U03](U04) {$(\La, 4)$};
            \draw [thick,>=latex']
            (U01.north) .. controls ($ (U01.north) + (0,2.5) $)
                and ($ (U04.north) + (0,2.5) $) .. (U04.north);
            \draw [thick,>=latex']
            (U02.north) .. controls ($ (U02.north) + (0,1.5) $)
                and ($ (U03.north) + (0,1.5) $) .. (U03.north);
        \end{tikzpicture}\vspace{-1.5em}
        \end{split}
    \end{equation*}

    \vspace{-1.25em}
    is not inclusion-free.
\end{block}}
\end{frame}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%o%%%%%%%%%%%%%%%%%%
\begin{frame} \frametitle{Main motivations}
Let $S_n$ be the set of permutations of size $n$.
\bigskip

\uncover<2->{%
There are several products on permutations, analogs of the shuffle
product of words. Among these, there are
\smallskip

@@

    \item the supershuffle \BIB{Vargas}{2014}.
\end{itemize}
\bigskip}

\uncover<3->{%
\begin{block}{Main questions}
    \begin{itemize}
        \item Combinatorial properties of ``square permutations''
        w.r.t. the supershuffle?
        \medskip

        \item Complexity of the recognition of ``square permutations''?
    \end{itemize}
\end{block}}
\end{frame}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{frame} \frametitle{Example: unshuffling binary words}
Let $B := \{\La, \Lb\}^*$ be the set of all binary words and consider
the coproduct
\begin{equation*}
    \Delta : \QQ[B] \to \QQ[B] \otimes \QQ[B]
\end{equation*}
on binary words defined by
\begin{equation*}
    \Delta(u) :=
    \sum_{P \sqcup Q = \{1, \dots, |u|\}}
    u_{|P} \otimes u_{|Q}.
\end{equation*}

\uncover<2->{%
\begin{block}{Example}\vspace{-1em}
\begin{equation*}
    \Delta(\Lb\La\La) =
    \epsilon \otimes \Lb\La\La + \Lb \otimes \La\La
    + 2 (\La \otimes \Lb\La) + 2 (\Lb\La \otimes \La) + \La\La \otimes \Lb
    + \Lb\La\La \otimes \epsilon
\end{equation*}
\end{block}}
\end{frame}

@@

\uncover<2->{%
\begin{block}{Example}
Since
\begin{equation*}
    \textcolor{RoyalBlue}{\Delta(\Lb\La\La)} =
    \epsilon \otimes \Lb\La\La + \Lb \otimes \La\La
    + \textcolor{Orange}{2} (\textcolor{BrickRed}{\La \otimes \Lb\La})
    + 2 (\Lb\La \otimes \La) + \La\La \otimes \Lb
    + \Lb\La\La \otimes \epsilon,
\end{equation*}
we observe that $\textcolor{BrickRed}{\La \otimes \Lb\La}$ has
multiplicity $\textcolor{Orange}{2}$ in
$\textcolor{RoyalBlue}{\Delta(\Lb\La\La)}$.
\bigskip

\uncover<3->{%
By duality, this implies that $\textcolor{RoyalBlue}{\Lb\La\La}$ has
multiplicity $\textcolor{Orange}{2}$ in
$\textcolor{BrickRed}{\La \shuffle \Lb\La}$. Indeed,
\begin{equation*}
    \textcolor{BrickRed}{\La \shuffle \Lb\La} =
    \La\Lb\La + \textcolor{Orange}{2}\, \textcolor{RoyalBlue}{\Lb\La\La}.
\end{equation*}}
\vspace{-1.5em}
\end{block}}
\end{frame}

@@

\uncover<2->{%
\begin{block}{Example}
\begin{equation*}
    \STD(82194) = 42153
\end{equation*}
\end{block}
\medskip}

\uncover<3->{%
\begin{block}{Definition}
    The \alert{unshuffling coproduct of permutations} is the
    coproduct on $\QQ[S]$ defined by
    \begin{equation*}
        \Delta(\pi) :=
        \sum_{P \sqcup Q = \{1, \dots, |\pi|\}}
        \STD\left(\pi_{|P}\right) \otimes \STD\left(\pi_{|Q}\right).
    \end{equation*}
\end{block}}

\uncover<4->{%
\begin{block}{Example}
\begin{equation*}
    \Delta(213) =
    \epsilon \otimes 213 + 2 (1 \otimes 12) +
    1 \otimes 21 + 2 (12 \otimes 1) + 21 \otimes 1
    + 213 \otimes \epsilon
\end{equation*}
\end{block}}
\end{frame}

@@

\uncover<2->{%
\begin{block}{Proposition}
    The permutations appearing in $\sigma \SHUFFLE \nu$ are exactly the
    one obtained by shuffling two words $u$ and $v$ respectively
    order-isomorphic to $\sigma$ and $\nu$.
\end{block}
\medskip}

\uncover<3->{%
\begin{block}{Example}\vspace{-1em}
\begin{equation*}\begin{split}
    12 \SHUFFLE 21 & =
    1243 + 1324 + 2 (1342) + 2 (1423) + 3 (1432) +
    2134 + 2 (2314) \\[.5em]
    & + 3 (2341) + 2413 + 2 (2431) + 2 (3124) + 3142 +
    3 (3214) + 2 (3241) \\[.5em]
    & + 3421 + 3 (4123) + 2 (4132) + 2 (4213) + 4231 + 4312
\end{split}\end{equation*}
\end{block}}
\end{frame}

@@

\uncover<2->{%
\begin{block}{Proposition}
    A permutation $\pi$ is a square iff $\pi$ can be obtained by
    shuffling two order-isomorphic words.
\end{block}
\medskip}

\uncover<3->{%
\begin{block}{Example}
    The permutation
    \begin{math}
        \pi :=
        \textcolor{RoyalBlue}{25}\textcolor{BrickRed}{16}
        \textcolor{RoyalBlue}{7}\textcolor{BrickRed}{83}
        \textcolor{RoyalBlue}{4}
    \end{math}
    is square since $\pi$ can be obtained by shuffling
    $\textcolor{BrickRed}{1683}$ and $\textcolor{RoyalBlue}{2574}$
    and
    \begin{math}
        \STD(\textcolor{BrickRed}{1683})
        = \STD(\textcolor{RoyalBlue}{2574})
        = 1342
    \end{math}.
\end{block}}
\end{frame}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{frame} \frametitle{Some properties of square permutations}
The first numbers of square permutations of size $2n$ are
\begin{equation*}
    1, 2, 20, 504, 21032, 1293418.
\end{equation*}

\uncover<2->{%
\begin{block}{Proposition}
    Let $\pi$ be a square permutation and $\sigma$ be a square root of
    $\pi$. Then,
    \begin{enumerate}
        \item $\widetilde{\pi}$ is a square and
        $\widetilde{\sigma}$ is one of its square roots;
        \smallskip

@@

        \item $\pi^{-1}$ is a square and $\sigma^{-1}$ is one of
        its square roots.
    \end{enumerate}
\end{block}
\medskip}
